---
title: "Multi-Lesion RECIST Progression and PFS/OS Modeling"
subtitle: "Pharmacometrics-inspired modeling in oncology (part 3)"
author: "Juliette Bouli-Mengue"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Motivation

Notebooks 1 and 2 modeled a single aggregate tumor size (SLD) per patient. In reality,
RECIST 1.1 (Eisenhauer et al., 2009; doi:10.1016/j.ejca.2008.10.026) criteria are based
on the **sum** of individually measured target lesions (up to 5 in total and up to 2 per
organ), each of which may respond differently to treatment even within the same patient.
This notebook implements the target-lesion component plus an assumed new-lesion process, and extends the framework by:

  - simulating multiple individual lesions per patient,
  - deriving total SLD as their sum,
  - detecting disease progression using RECIST 1.1 rules (rather than treating   progression as an abstract hazard, as in previous notebooks),
  - constructing Progression-Free Survival (PFS) as a composite time-to-event endpoint (the earliest of RECIST progression or death),
  - comparing PFS and OS on the same simulated patients.

```{r libraries}
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)

project_root <- Sys.getenv("TUMOR_PROJECT_ROOT", unset = "..")
source(file.path(project_root, "R", "config.R"))
source(file.path(project_root, "R", "recist.R"))
set_project_seed()
```

**Setup check.** Shared RECIST functions and the fixed seed ensure that lesion
generation, progression detection, and endpoint construction are reproducible without
copying business rules into the notebook.

## 2. Loading the simulated population

We reuse the simulated oncology population generated in notebook 1: each patient has
individual tumor growth parameters (`g`, `d`) representing their overall disease
trajectory. These patient-level parameters will serve as the basis for simulating
several individual lesions per patient in Section 3.

```{r load-data}
# Assumes patients_final.rds is saved in the same folder as this notebook
# (produced by notebook 1), a relative path, so the project stays portable across machines.
patients <- readRDS("patients_final.rds")

head(patients)
```

**Input check.** The same 200-patient population from Notebook 01 is reused. Survival
ground truth therefore stays fixed while the disease representation changes from one
aggregate burden to multiple lesions.

## 3. Simulating multiple lesions per patient

RECIST-based tumor burden is the **sum** of several individually measured target lesions (up to 5 in clinical practice), each of which may respond differently to
treatment even within the same patient. 
Rather than continuing to treat SLD as a single aggregate value per patient, we simulate 5 to 10 individual lesions per patient, each with its own growth (`g_lesion`) and regrowth (`d_lesion`) rate, and its own baseline size (`SLD0_lesion`). All derived from the patient-level parameters (`g`, `d`, `SLD0`) from notebook 1, with lesion-specific variation applied on top.

```{r create-lesions}
n_lesions_min <- 5
n_lesions_max <- 10

lesions <- patients %>%
  rowwise() %>%
  do({
    pt <- .
    n_lesions <- sample(n_lesions_min:n_lesions_max, 1)

    tibble(
      patient_id = pt$patient_id,
      lesion_id = 1:n_lesions,
      g_lesion = pt$g * runif(n_lesions, 0.7, 1.3),
      d_lesion = pt$d * runif(n_lesions, 0.7, 1.3),
      SLD0_lesion = (pt$SLD0 / n_lesions) * runif(n_lesions, 0.7, 1.3)
    )
  }) %>%
  ungroup()

head(lesions, 15)
cat("Number of lesions per patient,  summary:\n")
summary(table(lesions$patient_id))

lesions %>%
  group_by(patient_id) %>%
  summarise(n_lesions = n()) %>%
  summarise(min_lesions = min(n_lesions), max_lesions = max(n_lesions), mean_lesions = mean(n_lesions))
```

**Observation.** Each patient receives between five and ten lesions as specified. The
lesion parameters remain anchored to patient-level dynamics, while bounded variation
introduces within-patient heterogeneity without creating unrelated disease courses.

## 4. Simulating individual lesion trajectories and total tumor burden

Each lesion follows the same biexponential TGI equation used in notebook 1, but with its own lesion-specific parameters instead of the patient-level ones. 
We simulate scans every 6 weeks over 104 weeks, matching the follow-up duration used for the survival outcomes in notebook 1, then sum all lesions belonging to the same patient at each time point to obtain total SLD, exactly as RECIST defines it.

```{r create-trajectories}
assessment_times <- seq(0, 104, by = 6)

lesion_trajectories <- lesions %>%
  rowwise() %>%
  do({
    ls <- .
    tibble(
      patient_id = ls$patient_id,
      lesion_id = ls$lesion_id,
      time = assessment_times,
      diameter = ls$SLD0_lesion * (exp(-ls$g_lesion * assessment_times) + exp(ls$d_lesion * assessment_times) - 1)
    )
  }) %>%
  ungroup()

total_sld <- lesion_trajectories %>%
  group_by(patient_id, time) %>%
  summarise(SLD_total = sum(diameter), .groups = "drop")

head(total_sld, 12)

```

**Result.** Lesion diameters are aggregated at each assessment into one total SLD per
patient. The unit of the subsequent RECIST analysis is therefore explicit: progression
is evaluated on total target-lesion burden, not independently lesion by lesion.
## 5. Detecting RECIST progression

At each scan, we compare the current total SLD to the **nadir**, the lowest SLD value observed for that patient up to and including that time point. 
Progression is declared the first time both RECIST conditions are met simultaneously: at least a 20% increase
from nadir, and at least a 5mm absolute increase from nadir.

```{r detect-progression}
recist_data <- detect_progression(total_sld, "SLD_total")

head(recist_data, 12)
recist_data %>% filter(patient_id == 4) %>% select(time, SLD_total, nadir, relative_increase, absolute_increase, is_progression)
```

**Quality check.** The patient example exposes every component of the rule—the running
nadir, relative increase, absolute increase, and final flag—so the progression decision
can be audited instead of treated as a black box.

## 6. Extracting time to RECIST progression

For each patient, we identify the first time point at which RECIST progression is detected. Patients who never meet the progression criteria within the follow-up window
are treated as censored for progression at their last observed time point.

```{r extract-progression-times}
target_progression_times <- extract_progression_times(recist_data)

# New lesions are an independent RECIST progression criterion. They appear as a Poisson
# process (assumed constant rate, ~30% of patients over 104 weeks) and are only observed
# at the next scheduled scan. The earliest of target-lesion progression and new-lesion
# detection defines progression.
new_lesion_rate <- -log(0.70) / 104
new_lesion_times <- simulate_new_lesion_times(
  target_progression_times$patient_id, new_lesion_rate, assessment_times
)
progression_times <- add_new_lesion_progression(target_progression_times, new_lesion_times)

head(progression_times, 10)
table(progression_times$progression_cause)
cat("Number of patients with RECIST progression detected:", sum(progression_times$progression_event), "out of", nrow(progression_times), "\n")
progression_times %>% filter(progression_event == 0)
```

**Observation.** The extraction produces one progression record per patient and keeps
non-progressors as censored observations at their last scan. Patients without observed
progression are therefore not silently discarded. The `progression_cause` column shows
how many progressions come from target-lesion growth versus a new lesion. The new-lesion
rate is an illustrative assumption, not an estimate from clinical data.

## 7. Constructing PFS as a composite endpoint

Progression-Free Survival (PFS) is a composite time-to-event endpoint: a patient experiences a PFS event at the earliest of RECIST progression or death, whichever occurs first. We combine the RECIST progression times from Section 6 with the OS data from notebook 1 to construct PFS.

```{r construct-pfs}
# Non-informative loss to follow-up (assumed rate 0.002/week, ~19% of patients over
# 104 weeks), applied identically to PFS and OS. It censors patients before their
# event, which is how real trials avoid a near-100% event rate.
dropout_rate <- 0.002
dropout_time <- setNames(rexp(nrow(patients), rate = dropout_rate), patients$patient_id)

pfs_data <- patients %>%
  select(patient_id, os_time, os_event) %>%
  left_join(progression_times, by = "patient_id") %>%
  mutate(
    progression_time_filled = if_else(progression_event == 1, progression_time, last_observed_time),
    pfs_time_true = pmin(progression_time_filled, os_time),
    pfs_event_true = as.integer(progression_event == 1 | os_event == 1),
    dropout = dropout_time[as.character(patient_id)],
    pfs_time = pmin(pfs_time_true, dropout),
    pfs_event = as.integer(pfs_event_true == 1 & pfs_time_true <= dropout),
    os_event = as.integer(os_event == 1 & os_time <= dropout),
    os_time = pmin(os_time, dropout)
  )

head(pfs_data, 10)
```

**Endpoint decision.** PFS is the earliest of detected RECIST progression (target or new
lesion) and death, censored at loss to follow-up. The component times and event indicators remain in the table, preserving traceability
for every composite endpoint.

## 8. Comparing PFS and OS

As expected clinically, PFS events typically occur earlier than OS events, since disease progression is usually detected before death. We visualize both endpoints on the same population using Kaplan-Meier curves.

```{r compare-pfs-os}
comparison_data <- bind_rows(
  pfs_data %>% transmute(patient_id, time = pfs_time, event = pfs_event, endpoint = "PFS"),
  pfs_data %>% transmute(patient_id, time = os_time, event = os_event, endpoint = "OS")
)

km_comparison <- survfit(Surv(time, event) ~ endpoint, data = comparison_data)

ggsurvplot(km_comparison, data = comparison_data, risk.table = TRUE, pval = FALSE,
           xlab = "Time (weeks)", ylab = "Survival probability",
           title = "PFS vs. OS on the same simulated population",
           legend.title = "Endpoint")

comparison_data %>%
  group_by(endpoint) %>%
  summarise(median_time = median(time), event_rate = mean(event))
```

**Result.** Median PFS is **30 weeks**, earlier than median OS at **46.2 weeks**. The PFS
event rate is **94.5%**, compared with **72.5%** for OS. This ordering is coherent with
the composite definition: progression can occur before death and therefore adds earlier
events. Adding new-lesion progression and loss to follow-up moved these figures from the
earlier target-lesion-only version (36 vs. 52.3 weeks, 98.5% vs. 77%), but the PFS event
rate remains close to saturation. The cause is the notebook 1 regrowth model: the
`exp(d*t)` term makes almost every simulated tumor progress within 104 weeks, and
censoring can only remove a fraction of events. A lower event rate would require a
different regrowth distribution in notebook 1, not another endpoint construction. This
is therefore a stress-test scenario, not a realistic clinical-trial event rate, and it
limits discrimination late in follow-up.

## 9. Conclusion, limitations, and decisions for Notebook 04

This notebook extends the tumor dynamics framework from notebooks 1 and 2 to a
multi-lesion structure, closer to how RECIST is actually applied in clinical trials:
total tumor burden as the sum of individually simulated lesions, progression detected
via explicit RECIST 1.1 rules, and PFS constructed as a genuine composite endpoint
rather than an independently simulated hazard.

**Decisions carried forward:**

1. Reuse the same RECIST functions for treated and untreated arms.
2. Evaluate progression independently in each counterfactual arm because each arm has
   its own nadir and tumor trajectory.
3. Combine this multi-lesion structure with the ADC mechanism and repeated dosing from
   Notebook 02.
4. Compare both PFS and OS, while expecting the modeled treatment effect to appear more
   directly in PFS.

**Limitations, stated explicitly:**

- Lesions are simulated as varying independently around the patient-level `g` and `d`
  from notebook 1, rather than being generated from a shared mechanistic process; this
  is a simplification of true inter-lesion heterogeneity (e.g., differential drug
  penetration or clonal resistance across metastatic sites).
- Scans continue at fixed 6-week intervals throughout the full follow-up window
  regardless of death, which is not clinically realistic (a deceased patient has no
  further scans); this does not bias `pfs_time` itself, since `pmin()` always resolves
  to the earlier of progression or death, but it is a modeling simplification worth
  noting.
- New lesion appearance is modeled as a homogeneous Poisson process with an assumed
  rate (~30% of patients over 104 weeks) and is not linked to tumor burden, drug
  exposure, or the patient's growth parameters. Non-target and non-measurable disease
  are not modeled, so this remains a simplified implementation, not complete RECIST 1.1
  adjudication.
- Loss to follow-up is assumed non-informative (constant rate, independent of disease
  course); informative dropout would bias PFS and OS.
