---
title: "Does the ADC Improve PFS and OS? A Multi-Lesion RECIST Comparison"
subtitle: "Pharmacometrics-inspired modeling in oncology (part 4)"
author: "Juliette Bouli-Mengue"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## 1. Motivation

Notebook 3 introduced multi-lesion RECIST progression and PFS as a composite endpoint,
using the biexponential growth/regrowth model from notebook 1 (no drug mechanism).
Notebook 2 introduced an explicit ADC pharmacokinetic/pharmacodynamic mechanism, but
worked with a single aggregate tumor size and a simplified landmark survival analysis.

This notebook combines both: individual lesions (as in notebook 3) are driven by the
ADC PK/PD mechanism (as in notebook 2), allowing us to compare PFS and OS between a
treated (ADC) arm and an untreated counterfactual arm, using the same RECIST-based
progression detection built in notebook 3.

**Two compatibility choices, made explicitly:**

- **Time scale**: this notebook works in **days** (as in notebook 2), extended to 728
  days (~104 weeks) to allow enough follow-up time for progression and death events, as
  in notebook 3.
- **Growth model**: lesions here follow a growth-minus-drug-effect structure
  (`dSLD/dt = d*SLD - Effect(t)*SLD`). The `d` parameter from notebook 1 represents
  intrinsic regrowth and is therefore used as the untreated growth rate. The `g`
  parameter represented treatment-driven regression in notebook 1 and is not reused
  as natural growth here.
  
```{r libraries}
library(dplyr)
library(ggplot2)
library(survival)
library(survminer)

project_root <- Sys.getenv("TUMOR_PROJECT_ROOT", unset = "..")
source(file.path(project_root, "R", "config.R"))
source(file.path(project_root, "R", "pk_pd.R"))
source(file.path(project_root, "R", "recist.R"))
set_project_seed()
```  

**Setup check.** PK/PD and RECIST logic are imported from the project modules used by
the earlier notebooks. This final notebook combines those validated components rather
than redefining them locally.

## 2. Loading the simulated population

We reuse the simulated oncology population generated in notebook 1: each patient has
individual tumor growth parameters (`g`, `d`) representing their overall disease
trajectory. These patient-level parameters will serve as the basis for simulating
several individual lesions per patient in Section 3.

```{r load-data}
# Assumes patients_final.rds is saved in the same folder as this notebook
# (produced by notebook 1), a relative path, so the project stays portable across machines.
patients <- readRDS("patients_final.rds")

head(patients)
```

**Input check.** The common population is loaded unchanged, preserving patient-level
heterogeneity and keeping the treated/no-treatment comparison paired at the level of
the underlying disease parameters.

## 3. Simulating multiple lesions per patient

As in notebook 3, each patient has 5 to 10 individual lesions. Each lesion carries an
intrinsic regrowth parameter (`d_lesion`) inherited from the patient-level `d`, with
lesion-specific variability. This prevents relapse from being attributed solely to
loss of drug exposure.

```{r create-lesions}
n_lesions_min <- 5
n_lesions_max <- 10

lesions <- patients %>%
  rowwise() %>%
  do({
    pt <- .
    n_lesions <- sample(n_lesions_min:n_lesions_max, 1)

    tibble(
      patient_id = pt$patient_id,
      lesion_id = 1:n_lesions,
      d_lesion = pt$d * runif(n_lesions, 0.7, 1.3),
      SLD0_lesion = (pt$SLD0 / n_lesions) * runif(n_lesions, 0.7, 1.3)
    )
  }) %>%
  ungroup()

head(lesions, 10)
```

**Observation.** The lesion table provides the common starting state for both arms.
Only the treatment mechanism will differ; lesion count, baseline burden, and intrinsic
growth are shared within patient.

## 4. Applying PK/PD: repeated dosing, one drug exposure profile per patient

Real ADC regimens involve repeated administration (typically every 21 days), not a
single dose. We model plasma concentration using the superposition principle: total
concentration at any time is the sum of contributions from all doses already
administered, each decaying independently from its own injection time. This produces
the characteristic sawtooth accumulation pattern of repeated dosing. For internal
consistency, `EC50 = 1` and the population elimination rate are retained from notebook
2. `Emax = 0.003` is deliberately lower than notebook 2's single-dose value of `0.05`:
repeated q3w administration sustains exposure, so copying the single-dose `Emax`
produces near-complete, unrealistic progression suppression. This is an empirical
scenario calibration, not a clinically estimated ADC parameter.

```{r pk-pd-setup}
lesions <- lesions %>%
  mutate(d_daily = d_lesion / 7)

dose <- 5     # mg/kg
V    <- 3     # L/kg
EC50 <- 1
Emax <- 0.003  # lower than notebook 2 because exposure is sustained by repeated dosing

patients <- patients %>%
  mutate(kel_i = rlnorm(n(), meanlog = log(0.05), sdlog = 0.20))

time_grid <- seq(0, 728, by = 7)  # days, weekly scans over ~104 weeks
dosing_interval <- 21  # days, typical q3w ADC regimen
dosing_times <- seq(0, 728, by = dosing_interval)

patient_pkpd <- patients %>%
  rowwise() %>%
  do({
    pt <- .
    conc <- repeated_dose_concentration(time_grid, pt$kel_i, dosing_times, dose, V)
    inhibition <- emax_effect(conc, Emax, EC50)
    tibble(patient_id = pt$patient_id, time = time_grid, concentration = conc, inhibition = inhibition)
  }) %>%
  ungroup()

ggplot(patient_pkpd %>% filter(patient_id == 1), aes(x = time, y = concentration)) +
  geom_line() +
  labs(title = "ADC concentration with repeated q3w dosing (patient 1)",
       x = "Time (days)", y = "ADC concentration") +
  theme_minimal()


head(patients %>% select(patient_id, kel_i))
```

**PK/PD observation.** Superposition produces the expected q3w sawtooth exposure while
patient-specific elimination rates preserve PK heterogeneity. Computing exposure once
per patient guarantees that all lesions belonging to that patient receive the same
systemic concentration profile.
## 5. Simulating each lesion under repeated-dosing treatment and the no-drug counterfactual

Each lesion now uses its patient's repeated-dosing inhibition profile (Section 4, computed once per patient) rather than a single-dose profile computed per lesion. As before, we simulate two trajectories per lesion: one under ADC treatment, and one no-drug counterfactual.

```{r simulate-lesion-trajectories}
lesion_trajectories <- data.frame()

for (i in 1:nrow(lesions)) {
  pid    <- lesions$patient_id[i]
  d_i    <- lesions$d_daily[i]
  SLD0_i <- lesions$SLD0_lesion[i]

  inhibition <- patient_pkpd$inhibition[patient_pkpd$patient_id == pid]

  SLD_adc     <- numeric(length(time_grid)); SLD_adc[1] <- SLD0_i
  SLD_no_drug <- numeric(length(time_grid)); SLD_no_drug[1] <- SLD0_i

  for (t in 2:length(time_grid)) {
    dt <- time_grid[t] - time_grid[t - 1]
    SLD_adc[t] <- SLD_adc[t - 1] + dt *
      (d_i * SLD_adc[t - 1] - inhibition[t - 1] * SLD_adc[t - 1])
    SLD_no_drug[t] <- SLD_no_drug[t - 1] + dt *
      (d_i * SLD_no_drug[t - 1])

    # Euler's method can become non-physical under extreme parameter combinations.
    if (!is.finite(SLD_adc[t]) || SLD_adc[t] < 0) {
      stop("Invalid ADC tumor size; reduce the integration step", call. = FALSE)
    }
  }

  lesion_trajectories <- bind_rows(
    lesion_trajectories,
    data.frame(patient_id = pid, lesion_id = lesions$lesion_id[i], time = time_grid,
               SLD_adc = SLD_adc, SLD_no_drug = SLD_no_drug)
  )
}

head(lesion_trajectories, 10)
```

**Result.** Each lesion has two aligned trajectories on the same time grid: repeated-dose
ADC and no-drug growth. This paired construction is retained through aggregation,
RECIST assessment, and survival simulation.
## 6. Total tumor burden per arm

We sum all lesions belonging to the same patient, separately for each arm (ADC and no-drug counterfactual), to obtain total SLD over time — the RECIST-based tumor burden metric used for progression detection in Section 7.

```{r total-sld-by-arm}
total_sld_by_arm <- lesion_trajectories %>%
  group_by(patient_id, time) %>%
  summarise(
    SLD_total_adc = sum(SLD_adc),
    SLD_total_no_drug = sum(SLD_no_drug),
    .groups = "drop"
  )

head(total_sld_by_arm, 10)
```

**Aggregation check.** The result contains one row per patient and time point with a
separate total SLD for each arm. No lesion-level information is used directly by the
RECIST function after this aggregation.
## 7. Detecting RECIST progression on both arms

We apply the same simplified target-lesion component of the RECIST 1.1 progression rule
from notebook 3 (Eisenhauer et al., 2009; doi:10.1016/j.ejca.2008.10.026): ≥20% increase
from nadir AND ≥5mm absolute increase. It is evaluated independently in each arm because
the arms represent different counterfactual disease courses for the same patient.

```{r detect-progression-both-arms}
recist_adc <- detect_progression(total_sld_by_arm, "SLD_total_adc")
recist_no_drug <- detect_progression(total_sld_by_arm, "SLD_total_no_drug")

head(recist_adc %>% select(patient_id, time, SLD_total_adc, nadir, is_progression), 10)
```

**Decision.** Each arm receives its own running nadir and progression assessment. Using
one arm’s nadir for the other would invalidate the comparison because treatment changes
both the minimum burden and the subsequent relative increase.
## 8. Constructing PFS and OS for both arms

For each arm, we extract the first time of RECIST progression. For OS, death is simulated
interval by interval from a hazard that depends on tumor burden at the **start** of that
interval. Thus an event at time `t` can only depend on information available before
`t`; the former use of tumor burden at day 728 to simulate death from day 0 has been
removed. This remains a prognostic association model, not a causal biological model.

```{r construct-pfs-both-arms}
prog_adc <- extract_progression_times(recist_adc)
prog_no_drug <- extract_progression_times(recist_no_drug)

# Recompute the treated trajectory used for OS with treatment stopped at the first
# detected progression. The pre-progression path, and therefore PFS, is unchanged.
progression_lookup <- setNames(prog_adc$progression_time, prog_adc$patient_id)
lesion_trajectories_os <- data.frame()

for (i in 1:nrow(lesions)) {
  pid <- lesions$patient_id[i]
  d_i <- lesions$d_daily[i]
  progression_time_i <- progression_lookup[as.character(pid)]
  if (is.na(progression_time_i)) progression_time_i <- Inf
  inhibition <- patient_pkpd$inhibition[patient_pkpd$patient_id == pid]

  SLD_adc_os <- numeric(length(time_grid))
  SLD_adc_os[1] <- lesions$SLD0_lesion[i]

  for (t in 2:length(time_grid)) {
    dt <- time_grid[t] - time_grid[t - 1]
    active_inhibition <- if (time_grid[t - 1] < progression_time_i) inhibition[t - 1] else 0
    SLD_adc_os[t] <- SLD_adc_os[t - 1] + dt *
      (d_i * SLD_adc_os[t - 1] - active_inhibition * SLD_adc_os[t - 1])
  }

  lesion_trajectories_os <- bind_rows(
    lesion_trajectories_os,
    data.frame(patient_id = pid, lesion_id = lesions$lesion_id[i],
               time = time_grid, SLD_adc_os = SLD_adc_os)
  )
}

total_sld_for_os <- total_sld_by_arm %>%
  select(patient_id, time, SLD_total_no_drug) %>%
  left_join(
    lesion_trajectories_os %>%
      group_by(patient_id, time) %>%
      summarise(SLD_total_adc = sum(SLD_adc_os), .groups = "drop"),
    by = c("patient_id", "time")
  )

baseline_hazard <- 0.002  # per day
beta_SLD <- 0.01
reference_sld <- mean(total_sld_for_os$SLD_total_adc[total_sld_for_os$time == 0])

simulate_time_varying_os <- function(data, sld_column, arm_label) {
  patient_paths <- split(data[order(data$patient_id, data$time), ], data$patient_id)

  bind_rows(lapply(patient_paths, function(path) {
    event_time <- max(path$time)
    event <- 0L

    for (j in 2:nrow(path)) {
      dt <- path$time[j] - path$time[j - 1]
      current_sld <- path[[sld_column]][j - 1]
      hazard <- baseline_hazard * exp(beta_SLD * (current_sld - reference_sld))
      event_probability <- 1 - exp(-hazard * dt)

      if (runif(1) < event_probability) {
        event_time <- path$time[j]
        event <- 1L
        break
      }
    }

    tibble(
      patient_id = path$patient_id[1], arm = arm_label,
      os_time = event_time, os_event = event
    )
  }))
}

os_data <- bind_rows(
  simulate_time_varying_os(total_sld_for_os, "SLD_total_adc", "ADC"),
  simulate_time_varying_os(total_sld_for_os, "SLD_total_no_drug", "No ADC")
)

pfs_data_adc <- prog_adc %>%
  left_join(os_data %>% filter(arm == "ADC") %>% select(patient_id, os_time, os_event), by = "patient_id") %>%
  mutate(
    progression_time_filled = if_else(progression_event == 1, progression_time, last_observed_time),
    pfs_time = pmin(progression_time_filled, os_time),
    pfs_event = as.integer(progression_event == 1 | os_event == 1),
    arm = "ADC"
  )

pfs_data_no_drug <- prog_no_drug %>%
  left_join(os_data %>% filter(arm == "No ADC") %>% select(patient_id, os_time, os_event), by = "patient_id") %>%
  mutate(
    progression_time_filled = if_else(progression_event == 1, progression_time, last_observed_time),
    pfs_time = pmin(progression_time_filled, os_time),
    pfs_event = as.integer(progression_event == 1 | os_event == 1),
    arm = "No ADC"
  )

pfs_data_combined <- bind_rows(pfs_data_adc, pfs_data_no_drug)

head(pfs_data_combined, 10)
pfs_data_combined %>% filter(patient_id %in% c(1, 2)) %>% select(patient_id, arm, progression_time, os_time, pfs_time, pfs_event)
```

**Endpoint check.** The combined table keeps progression, OS, and PFS components for
both arms. The displayed patient examples make the `min(progression, death)` logic
auditable before fitting Kaplan–Meier curves.
## 9. Comparing PFS and OS between ADC and No ADC arms

We now visualize whether the ADC treatment improves both PFS and OS relative to the no-drug counterfactual, using Kaplan-Meier curves for each endpoint.

```{r compare-arms-km}
km_pfs <- survfit(Surv(pfs_time, pfs_event) ~ arm, data = pfs_data_combined)

ggsurvplot(km_pfs, data = pfs_data_combined, risk.table = TRUE, pval = TRUE,
           xlab = "Time (days)", ylab = "Progression-free survival probability",
           title = "PFS: ADC vs. No ADC", legend.title = "Arm")

km_os <- survfit(Surv(os_time, os_event) ~ arm, data = pfs_data_combined)

ggsurvplot(km_os, data = pfs_data_combined, risk.table = TRUE, pval = TRUE,
           xlab = "Time (days)", ylab = "Overall survival probability",
           title = "OS: ADC vs. No ADC", legend.title = "Arm")

pfs_logrank <- survdiff(Surv(pfs_time, pfs_event) ~ arm, data = pfs_data_combined)
os_logrank <- survdiff(Surv(os_time, os_event) ~ arm, data = pfs_data_combined)
pfs_p_value <- pchisq(pfs_logrank$chisq, df = 1, lower.tail = FALSE)
os_p_value <- pchisq(os_logrank$chisq, df = 1, lower.tail = FALSE)
c(PFS_logrank_p = pfs_p_value, OS_logrank_p = os_p_value)

pfs_data_combined %>%
  group_by(arm) %>%
  summarise(median_pfs = median(pfs_time), median_os = median(os_time),
            pfs_event_rate = mean(pfs_event), os_event_rate = mean(os_event))
pfs_data_combined %>% filter(arm == "ADC") %>% summarise(never_progressed = mean(progression_event == 0))
```

**Result and interpretation.** The table, curves, and computed `PFS_logrank_p` and
`OS_logrank_p` above are the authoritative results for this version of the model.
Historical p-values are not retained in prose because correcting the integration step,
growth parameter, and OS mechanism changes the generated endpoints. Any observed
separation describes one seeded simulation and is not clinical evidence for a product.

## 10. General conclusion, limitations, and next decisions

This notebook connects the two previously separate strands of the project: the
multi-lesion RECIST framework (notebook 3) and the ADC PK/PD mechanism (notebook 2),
extended here to repeated q3w dosing via the pharmacokinetic superposition principle.
The corrected model estimates treatment contrasts for both PFS and OS without using a
future tumor measurement to generate earlier deaths. Their magnitude and significance
must be read from the executed notebook rather than from previously calibrated,
now-obsolete values.

Across the four notebooks, the project now provides one traceable chain from simulated
patient heterogeneity to longitudinal tumor dynamics, ADC exposure, lesion-level
response, RECIST progression, PFS, and OS. The main conclusion is methodological: the
framework connects these quantities coherently, while its calibrated parameters and
same-source counterfactuals remain unsuitable for clinical-effect claims.

**Decisions for a future iteration:**

1. Replace empirically calibrated PD and hazard parameters with literature-based or
   estimated values and document parameter uncertainty.
2. Extend the implemented stop-at-progression rule to model subsequent therapy, dose
   delays, reductions, and toxicity-driven discontinuations.
3. Add new-lesion progression and non-target disease to the RECIST implementation.
4. Run repeated simulation scenarios and report uncertainty intervals for treatment
   effects instead of a single seeded realization.
5. Validate the workflow on an independent clinical dataset before drawing any
   translational conclusion.

**Limitations, stated explicitly:**

- `Emax = 0.003` is an empirical repeated-dose scenario calibration and intentionally
  differs from notebook 2's single-dose value (`0.05`). Repeated q3w exposure made the
  larger value suppress progression almost completely. Neither value is a
  literature-derived or fitted ADC estimate.
- Euler updates explicitly multiply daily rates by the seven-day grid interval. A
  smaller integration grid or validated ODE solver remains preferable for numerical
  sensitivity analysis.
- OS uses a piecewise-constant, time-varying hazard based on tumor burden known at the
  beginning of each interval. Tumor burden is a simulated prognostic biomarker here;
  this association is not evidence that changing tumor size causally changes survival.
- For the OS trajectory, ADC effect stops at the first progression detected on the
  weekly grid. This avoids attributing continued post-progression benefit to the study
  treatment. Subsequent therapy, delayed discontinuation, dose modification, and
  toxicity-driven stopping are still not modeled.
- As in notebook 3, new lesion appearance is not modeled; only growth/shrinkage of
  existing target lesions is considered.
- As in notebook 2, all patients follow the same fixed q3w schedule; real-world
  dose modifications, delays, and discontinuations due to toxicity are not modeled.
